Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-08-07 chapters_altFormat/ch_results.tex
Put in Pawel and Vladimir edits

 The data taken has been compared with 7 commonly used transport theoretical models. These models took part in a large collaborative effort in order to standardize certain common elements of each models \cite{theoryComp1,theoryComp2}. To reduce uncertainties in the pion production mechanism, each model simulated nuclear matter in a box type simulation with fixed, well defined, initial conditions. Though these box calculations model an equilibrated system at finite density, which is  not replicated by experiment, the solutions  of  pion and $\Delta$ yields can be analytically calculated from the ansatz of statistical equilibrium, providing a good benchmark. This allowed for each model to systematically go through the numerical treatment and details in each model such as initialization of nuclei, stability of the model, numerical handing of pauli-blocking, etc. After satisfying these benchmark calculations, each model was used with what each author considered the most reliable input and model assumption.  These models were taken in their best configuration, without any prior knowledge of the experimental data. We then simulated the 4 systems measured in the \spirit TPC at an impact parameter of \SI{3}{\femto\metre} at \SI{270}{\MeVA} beam energy. While the numerical treatments of each model are reasonably similar, each model differs considerably in their treatment of pion and $\Delta$ dynamics. Some models contain modifications to how the pion behaves in nuclear matter, i.e. \emph{in-medium effects}, typically introduced by including a pion optical potential, which describes the pion scattering and absorption. Some models include the iso-scalar and the iso-vector delta potential, which are not very well constrained, but can be important in the production of pions \cite{baoan_deltapotential,inmedPionKo,inmedPionFeng}.

@@

As will be the re-occurring theme, there is a large variation in the predicted pion observables between theoretical models; this is greater than the variation between the stiff and soft symmetry energies within a particular model. However, the remedy for this discrepancy is for the authors of these models to include or better model some other physics that many of them have neglected, or it could be that the relevant physics is not constrained by other observables and needs to be constrain by further experiments and comparisons.

@@

Other unconstrained physics includes aspects of the mean field potentials for nucleons, pions, and deltas. A primary focus is on the isovector potentials, of which only nucleon potentials for have been varied in the calculations shown below. Some of the models include the momentum dependent nucleonic mean field potentials, and the mean field potentials for deltas and pions. Calculations have already identified that the overall pion yields are influenced by the mean field potentials for the deltas \cite{cozmaPC}, which may not be surprising since a similar sensitivity has govern the occurrence of for delta and pionic matter within neutron stars \cite{deltaNS,pionNS}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{ypt_sn132.png}
\caption{Rapidity and transverse momentum plot for $\pi^-$ and $\pi^+$ in the $\tin{132}{124}$ system. The measured half of the distribution describes the forward rapidities $y_o > 0$ and has no other extrapolations. Where as the reflected side shows the measured distribution assuming symmetry about $y_o = 0$.}
\label{fig:ptrap_sn132}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{ypt_sn108.png}
\caption{Rapidity and transverse momentum plot for $\pi^-$ and $\pi^+$ in the $\tin{108}{112}$ system. The measured half of the distribution describes the forward rapidities $y_o > 0$ and has no other extrapolations. Where as the reflected side shows the measured distribution assuming symmetry about $y_o = 0$.}
\label{fig:ptrap_sn108}
\end{figure}

@@

We have compared the total pion yields and ratios to the 7 common transport models for the systems measured. The table of the transport models are listed in Appendix~\ref{tb:pionyieldTheory}. Figure~\ref{fig:totalpiYield} shows the total pion yield for the four systems measured as compared with the models. The models plotted here are only the soft symmetry energy since the variation in model is much larger than the variation within a model between different symmetry energies. While some models make a reasonable approximation of a particular charge pion yields, no model reasonably predicts both.

@@

Figure~\ref{fig:totalpiRatio} shows the total single pion ratio, $R = Y(\pi^-)/Y(\pi^+)$, and the double ratio  $DR = R_{\tin{132}{124}}/R_{\tin{124}{112}}$, of the $\tin{132}{124}$ and $\tin{108}{112}$ systems. Here, the variation between models is much larger than the variation between the momentum dependence of the symmetry energy within a particular model. The symmetry energy variation (soft and stiff extremes) of two models -- $\chi$BUU and TuQMD -- is plotted as a wide band in the single ratio and in all models in the double ratio. The circular markers and band represents the total error bar in the data. Certainly it can be seen that the variation between symmetry energy extremes in the models, though small, still exists; as initially predicted \cite{baoan_piprod1,baoan_piprod2}, and the small error bars in the data would facilitate a detailed analysis to extract the high density behavior of the symmetry energy, if not for large disagreement between models.

@@

The FOPI collaboration has measured the total pion multiplicity resulting from ${}^{197}Au + {}^{197}Au$ collisions at several, higher, beam energies. In the $\tin{132}{124}$ data the asymmetry is $N/Z=1.56$ where as in the ${}^{197}Au + {}^{197}Au$ the $N/Z=1.493$. Since 4 beams were measured in this experiment, the $N/Z$ dependence was measured as seen in Figure~\ref{fig:totalpiRatio}. Here the dependence is fitted with a 2-nd order polynomial fit. To compare the pion ratio in the $\tin{132}{124}$ data with that of the lower N/Z in the FOPI experiment, we scaled by the pion ratio between $N/Z=1.56$ and 1.493 as given from the fitted polynomial line. Figure~\ref{fig:fopiPionRatio} shows the scaled pion ratio as compared with the FOPI pion ratio \cite{fopi}. The fitted function has the functional form of $p_1(E - p_2)^{-2}$ where $p_1$ and $p_2$ are free parameters, and only is meant to guide the eye. It is also worth mentioning that the pion ratio observed in the FOPI \SI{400}{\MeVA} setting was already considerably higher than what is expected from the $(N/Z)^2$ delta resonance model \cite{baoan_piprod1,baoan_piprod2}. It it worth mentioning that the two experiments differ considerably in their efficiencies as was mentioned before. The FOPI results heavily relied on extrapolations to lower pion energies which the \spirit TPC did not rely on \cite{fopi}. Without a systematic error calculation or discussion with the FOPI collaboration no further detail can be given except that the \spirit TPC data seems to be reasonable expectation and in general agreement with the FOPI data set. Scaling the other systems in the \spirit TPC data leads to almost the exact same scaled value of $\mathrm{Y}(\pi^-)/\mathrm{Y}(\pi^+) = \num{4.82}$ and would be redundant.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{pionSpectra_sn108_sum.png}
\caption{Pion spectra for the $\tin{108}{112}$ system and comparisons to 7 theoretical models. }
\label{fig:pionspectraSn108}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{pionSpectra_sn132_sum.png}
\caption{Pion spectra for the $\tin{132}{124}$ system and comparisons to 7 theoretical models. }
\label{fig:pionspectraSn132}
\end{figure}

Figures~\ref{fig:pionspectraSn132} and \ref{fig:pionspectraSn108} shows pion CM kinetic energy spectra for both the $\tin{132}{124}$ and $\tin{108}{112}$ systems respectively; corrected for efficiency and accounting for the solid angle of 4$\pi$. This data marks the first time pion spectral data has been measured at sub-threshold energies. In the pion spectra, we can see the effect of the Coulomb potential, which accelerates $\pi^+$ and decelerates $\pi^-$ particles, due to the positive charge of the nuclear medium. The low yield for the $\pi^+$ production at low energies is caused by the Coulomb barrier between protons which limits $\pi^+$ production at low energies. The Coulomb force has the largest effect on the spectra of the pions and will play an important role when observing the general shape of the pion spectral ratio, where it changes its shape completely.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{singleRatio108_log.png}
\caption{Pion spectral ratio $\tin{108}{112}$ system and comparisons to 7 theoretical models.}
\label{fig:SRsn108}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{singleRatio132_sum.png}
\caption{Pion spectral ratio $\tin{132}{124}$ system and comparisons to 7 theoretical models.}
\label{fig:SRsn132}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{doubleRatio_sum.png}
\caption{Comparing the double spectral ratio, $R_{\tin{132}{124}}/R_{\tin{108}{112}}$, to the 7 theoretical transport simulations.}
\label{fig:spectraDR}
\end{figure}

>>>>

 The data taken has been compared with 7 commonly used transport theoretical models. These models took part in a large collaborative effort in order to standardize certain common elements of each models \cite{theoryComp1,theoryComp2}. To reduce uncertainties in the pion production mechanism, each model simulated nuclear matter in a box type simulation with fixed, well defined, initial conditions. Though these box calculations model an equilibrated system at finite density, which is  not replicated by experiment, the solutions  of  pion and $\Delta$ yields can be analytically calculated from the ansatz of statistical equilibrium, providing a good benchmark. This allowed for each model to systematically go through the numerical treatment and details in each model such as initialization of nuclei, stability of the model, numerical handing of Pauli-blocking, etc. After satisfying these benchmark calculations, each model was used with what each author considered the most reliable input and model assumption.  These models were taken in their best configuration, without any prior knowledge of the experimental data. We then simulated the 4 systems measured in the \spirit TPC at an impact parameter of \SI{3}{\femto\metre} at \SI{270}{\MeVA} beam energy. While the numerical treatments of each model are reasonably similar, each model differs considerably in their treatment of pion and $\Delta$ dynamics. Some models contain modifications to how the pion behaves in nuclear matter, i.e. \emph{in-medium effects}, typically introduced by including a pion optical potential, which describes the pion scattering and absorption. Some models include the isoscalar and the isovector delta potential, which are not very well constrained, but can be important in the production of pions \cite{baoan_deltapotential,inmedPionKo,inmedPionFeng}.

@@

As will be the re-occurring theme, there is a large variation in the predicted pion observables between theoretical models; this is greater than the variation between the stiff and soft symmetry energies within a particular model. However, the remedy for this discrepancy is for the authors of these models to include or better model some other physics that many of them have neglected, or it could be that the relevant physics is not constrained by other observables and needs to be constrained by further experiments and comparisons.

@@

Other unconstrained physics includes aspects of the mean field potentials for nucleons, pions, and deltas. A primary focus is on the isovector potentials, of which only nucleon potentials has been varied in the calculations shown below. Some of the models include the momentum dependent nucleonic mean field potentials, and the mean field potentials for deltas and pions. Calculations have already identified that the overall pion yields are influenced by the mean field potentials for the deltas \cite{cozmaPC}, which may not be surprising since a similar sensitivity has govern the occurrence of delta and pionic matter within neutron stars \cite{deltaNS,pionNS}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{ypt_sn132.png}
\caption{Rapidity and transverse momentum plot for $\pi^-$ and $\pi^+$ in the $\tin{132}{124}$ system. The measured half of the distribution describes the forward rapidities $y_o > 0$ and has no other extrapolations. whereas the reflected side shows the measured distribution assuming symmetry about $y_o = 0$.}
\label{fig:ptrap_sn132}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{ypt_sn108.png}
\caption{Rapidity and transverse momentum plot for $\pi^-$ and $\pi^+$ in the $\tin{108}{112}$ system. The measured half of the distribution describes the forward rapidities $y_o > 0$ and has no other extrapolations. whereas the reflected side shows the measured distribution assuming symmetry about $y_o = 0$.}
\label{fig:ptrap_sn108}
\end{figure}

@@

We have compared the total pion yields and ratios to the 7 common transport models for the systems measured. The table of the transport models are listed in Appendix~\ref{tb:pionyieldTheory}. Figure~\ref{fig:totalpiYield} show the total pion yield for the four systems measured as compared with the models. The models plotted here are only the soft symmetry energy since the variation in model is much larger than the variation within a model between different symmetry energies. While some models make a reasonable approximation of a particular charge pion yields, no model reasonably predicts both.

@@

Figure~\ref{fig:totalpiRatio} shows the total single pion ratio, $R = Y(\pi^-)/Y(\pi^+)$, and the double ratio  $DR = R_{\tin{132}{124}}/R_{\tin{124}{112}}$, of the $\tin{132}{124}$ and $\tin{108}{112}$ systems. Here, the variation between models is much larger than the variation between the momentum dependence of the symmetry energy within a particular model. The symmetry energy variation (soft and stiff extremes) of two models -- $\chi$BUU and TuQMD -- is plotted as a wide band in the single ratio and in all models in the double ratio. The circular markers and band represent the total error bar in the data. Certainly it can be seen that the variation between symmetry energy extremes in the models, though small, still exists; as initially predicted \cite{baoan_piprod1,baoan_piprod2}, and the small error bars in the data would facilitate a detailed analysis to extract the high density behavior of the symmetry energy, if not for large disagreement between models.

@@

The FOPI collaboration has measured the total pion multiplicity resulting from ${}^{197}Au + {}^{197}Au$ collisions at several, higher, beam energies. In the $\tin{132}{124}$ data the asymmetry is $N/Z=1.56$ whereas in the ${}^{197}Au + {}^{197}Au$ the $N/Z=1.493$. Since 4 beams were measured in this experiment, the $N/Z$ dependence was measured as seen in Figure~\ref{fig:totalpiRatio}. Here the dependence is fitted with a 2-nd order polynomial fit. To compare the pion ratio in the $\tin{132}{124}$ data with that of the lower N/Z in the FOPI experiment, we scaled by the pion ratio between $N/Z=1.56$ and 1.493 as given from the fitted polynomial line. Figure~\ref{fig:fopiPionRatio} shows the scaled pion ratio as compared with the FOPI pion ratio \cite{fopi}. The fitted function has the functional form of $p_1(E - p_2)^{-2}$ where $p_1$ and $p_2$ are free parameters, and only is meant to guide the eye. It is also worth mentioning that the pion ratio observed in the FOPI \SI{400}{\MeVA} setting was already considerably higher than what is expected from the $(N/Z)^2$ delta resonance model \cite{baoan_piprod1,baoan_piprod2}. It it worth mentioning that the two experiments differ considerably in their efficiencies as was mentioned before. The FOPI results heavily relied on extrapolations to lower pion energies which the \spirit TPC did not rely on \cite{fopi}. Without a systematic error calculation or discussion with the FOPI collaboration no further detail can be given except that the \spirit TPC data seems to be reasonable expectation and in general agreement with the FOPI data set. Scaling the other systems in the \spirit TPC data leads to almost the exact same scaled value of $\mathrm{Y}(\pi^-)/\mathrm{Y}(\pi^+) = \num{4.82}$ and would be redundant.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{pionSpectra_sn108_sum.png}
\caption{Pion spectra for the $\tin{108}{112}$ system and comparisons to 7 theoretical models. }
\label{fig:pionspectraSn108}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{pionSpectra_sn132_sum.png}
\caption{Pion spectra for the $\tin{132}{124}$ system and comparisons to 7 theoretical models. }
\label{fig:pionspectraSn132}
\end{figure}

Figures~\ref{fig:pionspectraSn132} and \ref{fig:pionspectraSn108} show pion CM kinetic energy spectra for both the $\tin{132}{124}$ and $\tin{108}{112}$ systems respectively; corrected for efficiency and accounting for the solid angle of 4$\pi$. This data marks the first time pion spectral data has been measured at sub-threshold energies. In the pion spectra, we can see the effect of the Coulomb potential, which accelerates $\pi^+$ and decelerates $\pi^-$ particles, due to the positive charge of the nuclear medium. The low yield for the $\pi^+$ production at low energies is caused by the Coulomb barrier between protons which limits $\pi^+$ production at low energies. The Coulomb force has the largest effect on the spectra of the pions and will play an important role when observing the general shape of the pion spectral ratio, where it changes its shape completely.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{singleRatio108_log.png}
\caption{Pion spectral ratio $\tin{108}{112}$ system and comparisons to 7 theoretical models.}
\label{fig:SRsn108}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{singleRatio132_sum.png}
\caption{Pion spectral ratio $\tin{132}{124}$ system and comparisons to 7 theoretical models.}
\label{fig:SRsn132}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{doubleRatio_sum.png}
\caption{Comparing the double spectral ratio, $R_{\tin{132}{124}}/R_{\tin{108}{112}}$, to the 7 theoretical transport simulations.}
\label{fig:spectraDR}
\end{figure}

### 2020-08-07 chapters_altFormat/ch_summary.tex
Put in Pawel and Vladimir edits

The need for precise experimental constraints on the symmetry energy at high densities is paramount to understanding exotic, dense matter, such as neutron stars. Direct observation of these objects provides information about the total equation of state, which is primarily sensitive to the pressure in neutron matter. Only by laboratory measurements can one control the asymmetry of matter and isolate its dependence on the isospin asymmetry $\delta =(\rho_n - \rho_p)/\rho$. This information is critical to understanding the symmetry energy and associated isovector mean field potentials. These mean field potential contribute strongly to the chemical potentials, which control the phase transitions and the internal structure within stars. Heavy ion collisions provide the only laboratory environment to study these issues at the suprasaturation densities relevant to the neutron star interior.

@@

  %With this wealth of new data that we have obtained, we now begin to work with theorists to understand what new experimental pion observable is sensitive to which theoretical assumption. We will are exploring how the isoscalar and isovector $\Delta$ potential plays an important role in the pion production spectra and emission rates. 
  Also for the first time, the pion spectra was measured. The spectral ratio appears to provide information that is relevant to this question also providing much needed information about the dynamics of pions, even to very low pion energies. Of particular importance is the sensitivity of calculations to the high energy tails of the pion spectra. This is a promising observable since high energy pions  exit the nuclear matter sooner and are less affected by re-scattering and adsorption-re-emission processes, which dilute information about the EOS at the highest density. We expect that high energy pions will be more sensitive to the prevailing conditions at high density, while low energy pions are more sensitive to the energy available for their emission, such as the Coulomb potential, the pion optical potentials, and the $\Delta$ potential.

\subsection{Outlook}
Theoretical collaboration, which has made considerable progress improving details of the calculation, will further improve in light of this new data. It is clear that a serious effort must be made to include effects such as the symmetry potentials of pions and deltas; which most codes do not. It also appears to be possible to make first constraints on the generally unknown $\Delta$ potential by fitting the pion yield data \cite{cozmaPC}. Making such a constraint would allow for the possibility to start making constraints on the density dependence of the symmetry energy. The interactions with theoretical community that has coalesced around this effort is very exciting, and we are seeing real progress in transport theory. It seems very possible that the theoretical efforts will converge and provide solid interpretations of these and other data much more quickly than they did in the case of the symmetric matter EoS.

>>>>

The need for precise experimental constraints on the symmetry energy at high densities is paramount to understanding exotic, dense matter, such as neutron stars. Direct observation of these objects provides information about the total equation of state, which is primarily sensitive to the pressure in neutron matter. Only by laboratory measurements can one control the asymmetry of matter and isolate its dependence on the isospin asymmetry $\delta =(\rho_n - \rho_p)/\rho$. This information is critical to understanding the symmetry energy and associated isovector mean field potentials. These mean field potentials contribute strongly to the chemical potentials, which control the phase transitions and the internal structure within stars. Heavy ion collisions provide the only laboratory environment to study these issues at the suprasaturation densities relevant to the neutron star interior.

@@

  %With this wealth of new data that we have obtained, we now begin to work with theorists to understand what new experimental pion observable is sensitive to which theoretical assumption. We will are exploring how the isoscalar and isovector $\Delta$ potential plays an important role in the pion production spectra and emission rates. 
  Also for the first time, the pion spectra were measured. The spectral ratio appears to provide information that is relevant to this question  about the dynamics of pions, even to very low pion energies. Of particular importance is the sensitivity of calculations to the high energy tails of the pion spectra. This is a promising observable since high energy pions  exit the nuclear matter sooner and are less affected by re-scattering and adsorption-re-emission processes, which dilute information about the EOS at the highest density. We expect that high energy pions will be more sensitive to the prevailing conditions at high density, while low energy pions are more sensitive to the energy available for their emission, such as the Coulomb potential, the pion optical potentials, and the $\Delta$ potential.

\subsection{Outlook}
Theoretical collaboration, which has made considerable progress improving details of the calculation, will further improve in light of this new data. It is clear that a serious effort must be made to include effects such as the symmetry potentials of pions and deltas; which most codes do not. It also appears to be possible to make first constraints on the generally unknown $\Delta$ potential by fitting the pion yield data \cite{cozmaPC}. Making such a constraint would allow for the possibility to start making constraints on the density dependence of the symmetry energy. The interactions with theoretical community that has coalesced around this effort are very exciting, and we are seeing real progress in transport theory. It seems very possible that the theoretical efforts will converge and provide solid interpretations of these and other data much more quickly than they did in the case of the symmetric matter EoS.
